### 2022-12-16 lmcs/REBUTTAL-LETTER.md

  The authors: we have now commented about it and restructured this part of the paper.

- Since the orignal publication of this work, a few works on deadlock-freedom by typing that go beyond the languages that stem directly from the logical interpretation of session types have been published that should be included in the related work discussion. Please comment on the recent work by Rocha and Caires on shared state in a language inspired by differential linear logic:

  > Pedro Rocha, Luís Caires:
  > Propositions-as-types and shared state. Proc. ACM Program. Lang. 5(ICFP): 1-30 (2021)

  Please comment on the work by Qian, Kavvos and Birkedal using coexponentials:

  > Zesen Qian, G. A. Kavvos, Lars Birkedal:
  > Client-server sessions in linear logic. Proc. ACM Program. Lang. 5(ICFP): 1-31 (2021)

  The authors: thanks, we have now added these references and commented about the work.

### 2022-12-16 lmcs/REBUTTAL-LETTER.md

# Response to the reviewers

We thank the reviewers for their useful feedback and comments. We addressed their concerns in the paper and respond below detailing how we went about these changes.

# Review 1

- Did you find it straightforward to apply the concepts from PCP to PGV? What was challenging?

@@

  The authors: The main difficulty was due to the fact that PGV has to track starting & ending priorities, rather than just [whichever PCP stores], but we were aided in this by the work by Padovani & Novara. We added the following passage:

  > We cannot straightforwardly adapt the priority typing from PCP to PGV, as PGV adds higher-order functions. Instead, the priority typing for PGV follow the work by Padovani and Novara~\cite{padovaninovara15}.

- page 4: In your presentation of types, I wondered if there are any well-formedness conditions on priority bounds.

@@

  The authors: There are no well-formedness conditions, but there are particular lemmas that hold for them, which could be _taken as_ well-formedness conditions. In particular I believe the theorem is: for a sequent with lower bound p and upper bound q, either p < q, or p is top and q is bottom. This is relaxed somewhat by some extensions, in which case I believe we can also have "p is top and q is a concrete priority", but I don't recall to what extend that applies here. I believe it may have been due to recursion, in which case it wouldn't. We added the following passage:

  > We postulate that the only function types---and, consequently, sequents---that are inhabited in PGV are pure functions and functions $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ for which $\cs{p}<\cs{q}$.

- page 6: The paragraph on values and evaluation contexts is going to be clear only to a selected few. Please write some text that explains how contexts work (holes, notation for replacing a hole with a term, etc.), that values are a specialisation of terms, etc.

@@

  The authors:  We added the following passage:

  > Values are the subset of terms which cannot reduce further.
  > Evaluation contexts are one-hole term contexts, \ie, terms with exactly one hole, written $\tm{\hole}$. We write $\tm{\plug{E}{M}}$ for the evaluation context $\tm{E}$ with its hole replaced by the term $\tm{M}$. Evaluation contexts are specifically those one-hole term contexts under which term reduction can take place.
  > Thread contexts are a convenient way to lift the notion of evaluation contexts to threads. We write $\tm{\plug{\conf{F}}{M}}$ for the thread context $\tm{\conf{F}}$ with its hole replaced by the term $\tm{M}$.
  > Configuration contexts are one-hole configuration contexts, \ie, configurations with exactly one hole, written $\tm{\hole}$. Specifically, configuration contexts are those one-hole term contexts under which configuration reduction can take place. The definition for $\tm{\conf{G}}$ only gives the case in which the hole is in the left-most parallel process, \ie, it only defines $\tm{\ppar{\conf{G}}{\conf{C}}}$ and not $\tm{\ppar{\conf{C}}{\conf{G}}}$. The latter is not needed, as $\tm{\parallel}$ is symmetric under structural congruence, though it would be harmless to add. We write $\tm{\plug{\conf{G}}{\conf{C}}}$ for the evaluation context $\tm{\conf{G}}$ with its hole replaced by the term $\tm{\conf{C}}$.

- Another example is that you say that you "recover fork", but at this point it's not clear what fork was in GV.

@@

  The authors: At this point the reader has been told that "[GV] combines channel creation and thread spawning into a single operation, called ``fork'', which is related to the cut construct in CP." This effectively summarizes all one needs to know about fork in GV. The fact that it is recoverable is only interesting to the readers familiar with GV, or interested in the exact relation to GV, and we hope that any reader unfamiliar with GV can gleam fork's original semantics from our definition.

- page 20, case E-Close: I couldn't make sense of the derivation, where two subderivations seem to be organised vertically?.. Is that the right reading? Could you reorganise it to make it obvious?

  The authors: we have now fixed this derivation.

- page 22: Lemma 3.7 is fundamental and seems to be the key difference wrt PCP and Kobayashi's work. Is that the case? Is the proof trivial? Could you at least provide a sketch, or say at least _why_ it is trivial?

### 2022-12-16 lmcs/REBUTTAL-LETTER.md

  > Values are the subset of terms which cannot reduce further.
  > Evaluation contexts are one-hole term contexts, \ie, terms with exactly one hole, written $\tm{\hole}$. We write $\tm{\plug{E}{M}}$ for the evaluation context $\tm{E}$ with its hole replaced by the term $\tm{M}$. Evaluation contexts are specifically those one-hole term contexts under which term reduction can take place.
  > Thread contexts are a convenient way to lift the notion of evaluation contexts to threads. We write $\tm{\plug{\conf{F}}{M}}$ for the thread context $\tm{\conf{F}}$ with its hole replaced by the term $\tm{M}$.
  > Configuration contexts are one-hole configuration contexts, \ie, configurations with exactly one hole, written $\tm{\hole}$. Specifically, configuration contexts are those one-hole term contexts under which configuration reduction can take place. The definition for $\tm{\conf{G}}$ only gives the case in which the hole is in the left-most parallel process, \ie, it only defines $\tm{\ppar{\conf{G}}{\conf{C}}}$ and not $\tm{\ppar{\conf{C}}{\conf{G}}}$. The latter is not needed, as $\tm{\parallel}$ is symmetric under structural congruence, though it would be harmless to add. We write $\tm{\plug{\conf{G}}{\conf{C}}}$ for the evaluation context $\tm{\conf{G}}$ with its hole replaced by the term $\tm{\conf{C}}$.

- Another example is that you say that you "recover fork", but at this point it's not clear what fork was in GV.

  The authors: At this point the reader has been told that "[GV] combines channel creation and thread spawning into a single operation, called ``fork'', which is related to the cut construct in CP." This effectively summarizes all one needs to know about fork in GV. The fact that it is recoverable is only interesting to the readers familiar with GV, or interested in the exact relation to GV, and we hope that any reader unfamiliar with GV can gleam fork's original semantics from our definition.

- page 12: Here and in all other cases in the remainder, you write "= with \Delta" on top which usually stands for a definition. Is that what you mean? Don't you mean a simple "=" instead? Or perhaps you mean that "= with \Delta" stands for "which is defined as"? A clarification would be appreciated.

  The authors: we intended to use the "= with \Delta" because the syntactic sugar rules are indeed definitions in terms of the core typing rules; they are macros for what comes after the "= with \Delta".

@@

- Abstract: It seems rather unusual to include citations.

  The authors: it is not uncommon for references to be included in the abstract. We think it gives a more complete picture, while allowing the abstract to remain succinct. Without the references we would need to expand on several notions, which defeats the purpose of a succinct abstract.

## General comments

- Another technical comment I have is on the distinction between variables and channels (which the authors refer to as names). The latter are generated at runtime (via new), and are also gathered, in the environment. I believe that’s also the reason why the usual theorems (preservation and progress) are stated relative to a non-empty context. However, the formalization should guarantee that the respective terms are closed. It would also be helpful if different meta variables were used for variables and names. Currently, x, y, z are used for both.

  The authors: we use 'names' to refer to channel endpoints, not channels. Preservation results (lemma 3.3 and thm 3.1) is standard and stated on open terms. Term progress (lemma 3.7) is stated for terms which may contain free names but not variables. We agree that separating variable and channel contexts would make the theorem statement clearer however, it would complicate the type system and its meta-theory. Configuration progress (thm 3.2) is stated for closed configurations, which is standard. Results are labelled accordingly. We clarified this after lemma 3.7.

### 2022-12-16 lmcs/REBUTTAL-LETTER.md

- page 4: In your presentation of types, I wondered if there are any well-formedness conditions on priority bounds.

  The authors: There are no well-formedness conditions, but there are particular lemmas that hold for them, which could be _taken as_ well-formedness conditions. In particular I believe the theorem is: for a sequent with lower bound p and upper bound q, either p < q, or p is top and q is bottom. This is relaxed somewhat by some extensions, in which case I believe we can also have "p is top and q is a concrete priority", but I don't recall to what extend that applies here. I believe it may have been due to recursion, in which case it wouldn't. We added a passage to the relevant section (now page 5).

- page 6: The paragraph on values and evaluation contexts is going to be clear only to a selected few. Please write some text that explains how contexts work (holes, notation for replacing a hole with a term, etc.), that values are a specialisation of terms, etc.

  The authors: We added a passage to the relevant section (now page 7).

@@

  The authors: we have now fixed this derivation.

- page 22: Lemma 3.7 is fundamental and seems to be the key difference wrt PCP and Kobayashi's work. Is that the case? Is the proof trivial? Could you at least provide a sketch, or say at least _why_ it is trivial?

  The authors: It's not _trivial_ but it's _boring_, as it's precisely the proof you'd give for any lambda calculus. We should mention that, essentially, this is just progress for the simply-typed linear lambda calculus, with "M is ready" falling out as extra stuck cases. We added the following passage:

  > [Progress for PGV] deviates from progress for linear lambda calculus only in that terms may reduce to values or _ready terms_, where the definition of ready terms encompasses all terms whose reduction is struck on some constant K.

- pages 27-36: These sections are too dry. There's no explanation of what works as usual, what is different, how the translation from PCP to PGV works (it is not even really _defined_ before the proofs!), and what the reader should focus on. It almost reads like a file intended for a theorem prover. I suggest that you rewrite these sections by explaining these aspects, presenting selected rules and the translation with examples, and then putting what "works as usual" into an appendix.

  The authors: We have added a section at the beginning of section 4.6 explaining how the translation works in prose.

@@

  The authors: we have fixed this now.

- page 4: You use a different font than usual for the empty set symbol, so maybe write explicitly in parenthesis that it is what you mean when you use the symbol for the first time?

  The authors: The rationale for colouring terms is to communicate their syntactic sort: red for terms, blue for types, teal for priorities & priority constraints. This may not have been consistently applied.

- page 23: In case (4.b), if L' is a value like an abstraction or a pair, what do you mean by "the endpoint that it is blocked on"?

  The authors: We amended the definition of _ready terms_ to include the phrase "is blocked on" as an alias for "is ready to act on".

### 2022-12-16 lmcs/priority-gv.tex
Changes

\emph{Ready terms} are terms which perform communication actions, either by themselves, \eg creating a new channel or thread, or with another thread, \eg sending or receiving.
It is worth mentioning that the notion of readiness presented here is akin to {live} processes introduced by Caires and Pfenning \cite{cairespfenning10,DardhaP22}, and {poised} processes introduced by Pfenning and Griffith \cite{PfenningG15} and later used by Balzer \etal \cite{balzerpfenning17,balzertoninho19}. Ready processes like live/poised processes denote processes that are ready to communicate on their providing channel.
\begin{definition}[Ready Terms]%
  \label{def:pgv-ready-actions}
  A~term $\tm{L}$ is \emph{ready} if it is of the form $\tm{\plug{E}{M}}$, where $\tm{M}$ is of the form $\tm{\new}$, $\tm{\spawn\;N}$, $\tm{\link\;\pair{x}{y}}$, or $\tm{M}$ acts on $\tm{x}$. In the latter case, we say that $\tm{L}$ is \emph{ready to act on} $\tm{x}$.
\end{definition}

@@

@@

  \begin{theorem}[Progress, $\cred$]%
    \label{thm:pgv-closed-progress-confs}
    If $\cseq[\main]{\emptyenv}{\conf{C}}$ and $\tm{\conf{C}}$ is in canonical form, then either $\tm{\conf{C}}\cred\tm{\conf{D}}$ for some $\tm{\conf{D}}$; or $\tm{\conf{C}\equiv\conf{D}}$ for some $\tm{\conf{D}}$ in normal form.
  \end{theorem}
  % \begin{proof}[Sketch]
  %   Our proof follows that of Kobayashi~\cite[theorem 2]{kobayashi06}. We apply \cref{lem:pgv-open-progress-terms} to each thread. Either we obtain a reduction, or each child thread is ready and the main thread ready or a value. We pick the ready term $\tm{L}$ with the smallest priority bound. If $\tm{L}$ contains new, spawn, or a link, we apply \LabTirName{E-New}, \LabTirName{E-Spawn}, or \LabTirName{E-Link}. Otherwise, $\tm{L}$ must be ready on some $\tm{x_i}$. Linearity guarantees there is some thread  $\tm{L'}$ which acts on $\tm{x'_i}$. If $\tm{L'}$ is ready, priority typing guarantees it is ready on $\tm{x'_i}$, and we apply \LabTirName{E-Send} or \LabTirName{E-Close}. If $\tm{L'}$ is not ready, it must be the main thread returning a value. We move $\tm{L}$ into the $i^{\text{th}}$ position and repeat until we either find a reduction or reach normal form.
  % \end{proof}
  \input{prf/thm-pgv-closed-progress-confs}

>>>>

\emph{Ready terms} are terms which perform communication actions, either by themselves, \eg creating a new channel or thread, or with another thread, \eg sending or receiving.
It is worth mentioning that the notion of readiness presented here is akin to {live} processes introduced by Caires and Pfenning \cite{cairespfenning10,DardhaP22}, and {poised} processes introduced by Pfenning and Griffith \cite{PfenningG15} and later used by Balzer \etal \cite{balzerpfenning17,balzertoninho19}. Ready processes like live/poised processes denote processes that are ready to communicate on their providing channel.
\begin{definition}[Ready Terms]%
  \label{def:pgv-ready-actions}
  A~term $\tm{L}$ is \emph{ready} if it is of the form $\tm{\plug{E}{M}}$, where $\tm{M}$ is of the form $\tm{\new}$, $\tm{\spawn\;N}$, $\tm{\link\;\pair{x}{y}}$, or $\tm{M}$ acts on $\tm{x}$. In the latter case, we say that $\tm{L}$ is \emph{ready to act on} $\tm{x}$ or is \emph{blocked on}.
\end{definition}

@@

With ``$\ty{\Gamma}$ contains only session types'' we mean that for every $\tmty{x}{T}\in\ty{\Gamma}$, $\ty{T}$ is a session type, \ie, is of the form $\ty{S}$.

@@

  \begin{theorem}[Progress, $\cred$]%
    \label{thm:pgv-closed-progress-confs}
    If $\cseq[\main]{\emptyenv}{\conf{C}}$ and $\tm{\conf{C}}$ is in canonical form, then either $\tm{\conf{C}}\cred\tm{\conf{D}}$ for some $\tm{\conf{D}}$; or $\tm{\conf{C}\equiv\conf{D}}$ for some $\tm{\conf{D}}$ in normal form.
  \end{theorem}
  \input{prf/thm-pgv-closed-progress-confs}
